\bf Model Visualization Techniques for a Social Network Model

### 2017-08-30 paper.Rnw
just about done with changes from meeting

<<logweibull, fig.align='center', out.width = '.5\\linewidth', fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
library(reliaR)
x <- seq(-3,10,.01)
y <- dgumbel(x = seq(-3,10,.01), sigma = 1, mu=0)
qplot(x = x, y= y, geom = "line") + 
  annotate('text', x = 6, y = .25, 
        label = "g(u) = exp(-(u + exp(-u)))",size=5)  + 
  labs(x = "u", y = "g(u) (density)")
@

@@

$$p_{ij}(\boldsymbol{\beta}, x) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j))\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h))\right\}}$$

@@

<<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network. \\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
p1 <- plot(gof1, period = 1)
p2 <- plot(gof2, period = 1)
gridExtra::grid.arrange(p1, p2)
# sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
# [1] 0.154
# > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
# [1] 0.063

@@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", fig.width=8, fig.height=3>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 4, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior") + 
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', panel.background = element_rect(fill = NA, color = "grey30"))
alldataviz
@

@@

<<tableofmeans>>=
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
senM1est <- read.csv("data/congress/senateM1ests.csv")
senM2est <- read.csv("data/congress/senateM2ests.csv")
senM3est <- read.csv("data/congress/senateM3ests.csv")
senM1est$Model <- "M1"
senM2est$Model <- "M2"
senM3est$Model <- "M3"
senM1est$Sim <- 1:nrow(senM1est)
senM2est$Sim <- 1:nrow(senM2est)
senM3est$Sim <- 1:nrow(senM3est)
senM1est %>% gather(parameter, estimate, Rate:recip) %>% select(Sim, Model, parameter, estimate) -> senM1est2 
senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% select(Sim, Model, parameter, estimate) -> senM2est2 
senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% select(Sim, Model, parameter, estimate) -> senM3est2 
senateEsts <- rbind(senM1est2, senM2est2, senM3est2)
senateEsts$parameter <- as.factor(senateEsts$parameter)
levels(senateEsts$parameter) <- c("beta1", "beta3", "beta4", "alpha1", "alpha2", "alpha3", "beta2")
SSE <-
  senateEsts %>% group_by(Model, parameter) %>% 
  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
  mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
  select(Model, parameter, val) %>% 
  spread(Model, val) %>% arrange(as.character(parameter))
SFE <- simu2 %>% group_by(Model, parameter) %>% 
  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
   mutate(val = paste0(round(meanEst,3), " ", "(", round(sdEst,3), ")")) %>% 
  select(Model, parameter, val) %>% 
  spread(Model, val) %>% arrange(parameter)
@

\begin{table}
\begin{tabular}{l|ccc|ccc }
  \hline
  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
  \hline
  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
  \hline
\end{tabular}
\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network. Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
\end{table}

@@

ggplot(data = clintonSenate) + 
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= T, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  #theme_net() + 
  ThemeNet + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

@@

<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs. The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.width=8>>=

@@

ggplot(data = simu2) + 
  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
  facet_wrap(~parameter, scales = 'free') + 
  ThemeNoNet
@

@@

<<corplotsSenate>>=
senM4est <- read.csv("data/congress/senateM4ests.csv")
senM5est <- read.csv("data/congress/senateM5ests.csv")
names(senM4est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta5")
names(senM5est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta6")
senM4est$Model <- "M4"
senM5est$Model <- "M5"
senM4est$Sim <- 1:nrow(senM4est)
senM5est$Sim <- 1:nrow(senM5est)
senM4est2 <- senM4est %>% gather(parameter, estimate, alpha1:beta5) 
senM5est2 <- senM5est %>% gather(parameter, estimate, alpha1:beta6) 
senM45est <- rbind(senM4est2, senM5est2)
senM45est %>% spread(parameter, estimate)-> senM45est2
ggpairs(senM45est2, columns = 3:9, ggplot2::aes(colour=Model, alpha = .5))
@

@@

<<makinggif, fig.cap='A selection of images in the microstep animation. The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', fig.width=8, cache=TRUE, message = FALSE, warning = FALSE>>=
# Need old version of dplyr to get this to work: 
if (!require("dplyr050",character.only = TRUE)) {
      devtools::install_github("sctyner/dplyr050")
}

>>>>

<<logweibull, fig.align='center', out.width = '.99\\linewidth', fig.height = 2, fig.cap='The probability distribution function for the type 1 extreme value distribution, also known as the log-weibull or Gumbel distribution with location parameter $ \\mu = 0$ and scale parameter $ \\sigma = 0$.'>>=
library(reliaR)
x <- seq(-3,10,.01)
y <- dgumbel(x = seq(-3,10,.01), sigma = 1, mu=0)
qplot(x = x, y= y, geom = "line") + 
  annotate('text', x = 6, y = .25, 
        label = "g(u) = exp(-(u + exp(-u)))",size=5)  + 
  labs(x = "u", y = "density of g(u)")
@

@@

\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j))\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h))\right\}}
\end{equation}

@@

<<sienagof, fig.cap="Goodness of fit measures for two network statistics computed on the first simulated wave of a SAOM: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the network. \\st{The p-value shown on the x-axis is the proportion of observations simulated from the model that have observed statistic values as large or larger than the values observed in the data.}", cache = TRUE, message = FALSE, warning = FALSE, fig.height=5>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)

sims1 <- data.frame(gof1$`Period 1`$Simulations)
dat1 <- gof1$`Period 1`$Observations
dat1 %>% data.frame() %>% gather(outdegree, val) %>% 
  mutate(outdegree = parse_number(outdegree)-1) -> dat1
p1 <- sims1 %>% 
  mutate(sim = row_number()) %>% 
  gather(outdegree, val, X1:X9) %>% 
  mutate(outdegree = parse_number(outdegree) - 1) %>% 
  filter(outdegree <= 6) %>%  
  ggplot(aes(x = outdegree, y = val)) + 
  geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) + 
  geom_violin(aes(group = outdegree), bw = 2, fill = NA) + 
  geom_point(data = dat1, color = 'red') + 
  geom_line(data= dat1, color = 'red') + 
  geom_text(data = dat1, aes(label = val), hjust = -.5) + 
  labs(x = "Outdegree (p = 0.154)", y = "Statistic", 
       title = "Goodness-of-Fit: Outdegree distribution period 1") + 
  ThemeNoNet

sims2 <- data.frame(gof2$`Period 1`$Simulations)
dat2 <- gof2$`Period 1`$Observations
dat2 %>% data.frame() %>% gather(indegree, val) %>% 
  mutate(indegree = parse_number(indegree)-1) -> dat2
p2 <- sims2 %>% 
  mutate(sim = row_number()) %>% 
  gather(indegree, val, X1:X9) %>% 
  mutate(indegree = parse_number(indegree) - 1) %>% 
  filter(indegree <= 6) %>%  
  ggplot(aes(x = indegree, y = val)) + 
  geom_boxplot(aes(group = indegree), outlier.shape = "x", outlier.size = 5) + 
  geom_violin(aes(group = indegree), bw = 2, fill = NA) + 
  geom_point(data = dat2, color = 'red') + 
  geom_line(data= dat2, color = 'red') + 
  geom_text(data = dat2, aes(label = val), hjust = -.5) + 
  labs(x = "Indegree (p = 0.063)", y = "Statistic", 
       title = "Goodness-of-Fit: Indegree distribution period 1") + 
  ThemeNoNet

gridExtra::grid.arrange(p1, p2)
# sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
# [1] 0.154
# > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
# [1] 0.063

@@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper.", out.width='.99\\linewidth', fig.height=3>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) + 
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', panel.background = element_rect(fill = NA, color = "grey30"))
alldataviz
@

@@

<<tableofmeans>>=
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
senM1est <- read.csv("data/congress/senateM1ests.csv")
senM2est <- read.csv("data/congress/senateM2ests.csv")
senM3est <- read.csv("data/congress/senateM3ests.csv")
senM1est$Model <- "M1"
senM2est$Model <- "M2"
senM3est$Model <- "M3"
senM1est$Sim <- 1:nrow(senM1est)
senM2est$Sim <- 1:nrow(senM2est)
senM3est$Sim <- 1:nrow(senM3est)
senM1est %>% gather(parameter, estimate, Rate:recip) %>% select(Sim, Model, parameter, estimate) -> senM1est2 
senM2est %>% gather(parameter, estimate, Rate:jumpXTransTrip) %>% select(Sim, Model, parameter, estimate) -> senM2est2 
senM3est %>% gather(parameter, estimate, Rate:nbrDist2twice) %>% select(Sim, Model, parameter, estimate) -> senM3est2 
senateEsts <- rbind(senM1est2, senM2est2, senM3est2)
senateEsts$parameter <- as.factor(senateEsts$parameter)
levels(senateEsts$parameter) <- c("beta1", "beta3", "beta4", "alpha1", "alpha2", "alpha3", "beta2")
SSE <-
  senateEsts %>% group_by(Model, parameter) %>% 
  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
  mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
  select(Model, parameter, val) %>% 
  spread(Model, val) %>% arrange(as.character(parameter))
SFE <- simu2 %>% group_by(Model, parameter) %>% 
  summarize(meanEst = mean(estimate), sdEst = sd(estimate)) %>% 
   mutate(val = paste0(sprintf("%.3f",meanEst), " ", "(", sprintf("%.3f", sdEst), ")")) %>% 
  select(Model, parameter, val) %>% 
  spread(Model, val) %>% arrange(parameter)
@

\begin{table}
\begin{tabular}{l|ccc|ccc }
  \hline
  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
  \hline
  M1 & M2 & M3 & M1 & M2 & M3 \\
  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
  \hline
\end{tabular}
\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network. Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
\end{table}

@@

ggplot(data = clintonSenate) + 
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= T, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  #theme_net() + 
  ThemeNet + 
  scale_shape_manual(values = c(17,16)) + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

@@

<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs. The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=4>>=

@@

ggplot(data = simu2) + 
  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
  facet_wrap(~parameter, scales = 'free') + 
  ThemeNoNet + 
  theme(legend.position = 'bottom')
@

@@

<<corplotsSenate, eval = FALSE>>=
senM4est <- read.csv("data/congress/senateM4ests.csv")
senM5est <- read.csv("data/congress/senateM5ests.csv")
names(senM4est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta5")
names(senM5est) <- c("alpha1", "alpha2", "alpha3", "beta1", "beta2", "beta6")
senM4est$Model <- "M4"
senM5est$Model <- "M5"
senM4est$Sim <- 1:nrow(senM4est)
senM5est$Sim <- 1:nrow(senM5est)
senM4est2 <- senM4est %>% gather(parameter, estimate, alpha1:beta5) 
senM5est2 <- senM5est %>% gather(parameter, estimate, alpha1:beta6) 
senM45est <- rbind(senM4est2, senM5est2)
senM45est %>% spread(parameter, estimate)-> senM45est2
ggpairs(senM45est2, columns = 3:9, ggplot2::aes(colour=Model, alpha = .5))
@

@@

<<makinggif, fig.cap='A selection of images in the microstep animation. The selected edges and nodes are emphasized by changing the size and the color, then when edges are removed, they fade out, shrinking in size, while the nodes change color and shrink to blend in with the other nodes.', cache=TRUE, message = FALSE, warning = FALSE, fig.height=3>>=
# Need old version of dplyr to get this to work: 
if (!require("dplyr050",character.only = TRUE)) {
      devtools::install_github("sctyner/dplyr050")
}
